DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-10-26 DCMIP2016-Part1-Models.tex
- Near final production update

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.

@@

The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}). Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The prognostic horizontal winds are stored in the native Gnomonic local coordinate. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
the vertical discretization all variables are co-located. 
The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.

%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
%are available by means of simple switches. The formulation incorporates 
%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
%Conjugate Residual approach. The dynamics time step is adapted 
%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
%than 1). The physics time step is identical to the dynamics time step. 
% 
%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
%The horizontal spatial discretization is fully unstructured finite-volume using the 
%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
%In both the horizontal and the vertical discretization, all prognostic variables are 
%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
%
%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
%the form of a Rayleigh-type forcing to the prognostic variables.

%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
%efficiency and accuracy.

%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
%framework developed at ECMWF.

@@

The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.  The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator, applied after the completion of the physics time step.

\subsection{Icosahedral Non-hydrostatic model (ICON)}

The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.  Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.  In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.  The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.  Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.  Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.

@@

NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}. The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}. Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}. All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids. The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}. The Arakawa C-grid staggering is used vertically. A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended. The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}. The implicit time discretization is adopted for the vertically propagating one. A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.

@@

OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form. Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting. Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons). Height is the vertical coordinate, and a Lorenz vertical grid staggering is used. A unique feature of OLAM is that grid levels are horizontal and intersect topography. This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms. Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground. The OLAM cut-cell formulation conserves mass and momentum.

@@

The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.  The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new, wood2014inherently}.

@@

The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed, PL2007JCP} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.  The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.  Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.  This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).

\begin{figure}
\begin{center}
\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
\end{center}
\caption{(Left) A cubed-sphere grid.  (Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.  (Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
\end{figure}

@@

The icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.  These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained. For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle. By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.

@@

\subsubsection{Mass-based coordinates}

@@

>>>>

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A is built upon an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration (\S \ref{sec:grid_cs}), although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates (\S \ref{sec:FloatingLagrangian}).
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.  Tracer transport is sub-cycled relative to the hydrodynamics, using the spectral element method, with tracer mass as the prognostic variable.

@@

The CSU model is a finite-volume model using an optimized geodesic grid \citep{heikes1995numerical,heikes2013optimized} (\S \ref{sec:grid_geo}), with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order hyperviscosity operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer (\S \ref{sec:diffusion_csu}).  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model using a geodesic grid (\S \ref{sec:grid_geo}) and a floating vertical mass coordinate (\S \ref{sec:FloatingLagrangian}).  Although originally a hydrostatic model, it has been recently extended to solve the shallow-atmosphere non-hydrostatic Euler equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  Horizontal diffusion is included via a fourth-order hyperviscosity operator (\S \ref{sec:diffusion_dynamico}).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid (\S \ref{sec:grid_cs}) with a floating Lagrangian vertical coordinate. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). The non-hydrostatic formulation extends the hydrostatic model described in  \cite{lin2004vertically} by adding a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density.  The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}), although the prognostic horizontal winds are stored in the native Gnomonic local coordinate.  All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.  Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. Since divergence is effectively invisible to the solver, a 2D divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale (\S \ref{sec:diffusion_fv3}).  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

@@

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the non-hydrostatic Euler equations on an octahedral reduced Gaussian grid (\S \ref{sec:grid_oct}) with a height-based terrain-following vertical coordinate \citep{szmelter2010,smolarkiewiczetalJCP2016}. The horizontal spatial discretization uses the median-dual finite-volume approach, combined with a structured-grid finite-difference method in the vertical. In both the horizontal and 
the vertical discretization all variables are co-located. A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}). The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}.  The non-oscillatory (i.e.~monotonic) MPDATA 
also provides sufficient dissipation/diffusion to stabilize the model, so no other explicit filtering mechanism is required (\S \ref{sec:diffusion_fvm}). 
Note that the octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes \citep[][]{szmelter2010,kuehnleinJCP2012,deconinck2017atlas}.

@@

The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.  GEM solves the non-hydrostatic Euler equations on the Yin-Yang grid \citep{kageama2004yinyang} (\S \ref{sec:grid_yinyang}) with Arakawa C-grid staggering of prognostic variables.
The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type (\S \ref{sec:GEM_zeta})
and the vertical discretization is based on the Charney-Phillips grid (\S \ref{sec:VerticalStaggering}).
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.  Semi-Lagrangian advection is also used for tracer transport.
To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator, applied after the completion of the physics time step (\S \ref{sec:diffusion_gem}).

\subsection{ICOsahedral Non-hydrostatic model (ICON)}

The ICON model \citep{zangl2015icon} is a finite-volume model that solves the non-hydrostatic Euler equations in 2D vector invariant form on an icosahedral (triangular) grid (\S \ref{sec:grid_ico_tri}) with Arakawa C-grid staggering, and further utilizing a smoothed terrain-following height-based Lorenz vertical discretization.  Prognostic horizontal velocities are stored as normal wind components at the edge mid-points of full levels.  Prognostic vertical velocity is stored at the circumcenters of the triangles on half levels. The discretization employs a two time-level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves.  For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow-dependent Smagorinsky scheme with a background 4th-order Laplacian diffusion operator (\S \ref{sec:diffusion_icon}). For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which leads to local mass conservation and consistency with the air motion. Specifically, the average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. These numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

%Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps.

@@

The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} is a finite-volume model that solves the non-hydrostatic Euler equations using an Arakawa C-grid staggering on a centroidal Voronoi tessellation mesh (\S \ref{sec:grid_ccvt}), and the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.  In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.  Advection is nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.  The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.  Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.  Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, fourth-order hyperdiffusion, and 2D and 3D divergence damping operators (\S \ref{sec:diffusion_mpas}).

@@

NICAM is a finite-volume model that solves the non-hydrostatic Euler equations using a geodesic grid (\S \ref{sec:grid_geo}) optimized with spring dynamics using the method of \cite{tomita2002optimization}. A terrain-following height coordinate system is used in the vertical \citep{tomita2004new} with Lorenz staggering. Instead of temperature or potential temperature, total energy is prognosed following the method of \citep{satoh2002conservative, satoh2003conservative}. All prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell-integrated quantities to point values at cell centroids. The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}. For integration in time, a two-stage Runge-Kutta scheme is usually employed because of low computational cost, although a three-stage Runge-Kutta scheme \citep{wicker2002time} is available and recommended. The split-explicit time discretization is used for the horizontally propagating sound waves with the 3D divergence damping term \citep{skamarock1992stability} (\S \ref{sec:diffusion_nicam}). An implicit time discretization is adopted for the vertically propagating wave modes. A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.

@@

OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a finite-volume model that solves the deep-atmosphere non-hydrostatic Euler equations in momentum conservation form on a hexagonal Voronoi mesh (\S \ref{sec:grid_geo}) with Arakawa C-grid staggering.  The model supports optional local mesh refinement, which introduces some pentagons and heptagons to the grid.  Height is the vertical coordinate, and a Lorenz vertical grid staggering is used.  A unique feature of OLAM is that grid levels are horizontal and intersect topography (\S \ref{sec:CutCells}). This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms. Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground. The OLAM cut-cell formulation conserves mass and momentum.  Acoustic modes are solved explicitly in the horizontal using time splitting and a second-order Lax Wendroff method, and implicitly in the vertical.  Tracer transport is second-order in space and time using the scheme of \cite{miura2007upwind}, with consistent fluxes obtained by time averaging over the acoustic time steps.

%Acoustic modes are solved explicitly in the horizontal using a 2nd-order Adams-Bashforth method and implicitly in the vertical by means of time splitting.  Tracer transport is 2nd-order in space and time, using the scheme of \cite{crowley1968numerical} for obtaining consistent numerical fluxes, and Lax-Wendroff for time integration.

%, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form.  Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons).

@@

%Based upon the sections for the rest of the paper I would like to be
%able to ascertain the following properties for each model from this section: equation
%set, horizontal grid and discretization, vertical grid and discretization, temporal
%discretization, principal diffusion and stabilization mechanisms and transport
%scheme.

%The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a finite-volume model that solves the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with a floating Lagrangian vertical coordinate.

The Tempest model \citep{ullrich2014global, guerra2016high} is an experimental testbed for high-performance numerical methods that solves the non-hydrostatic Euler equations on a cubed-sphere grid (\S \ref{sec:grid_cs}) using a horizontally co-located spectral element discretization.  In the vertical Tempest uses an Eulerian finite-volume discretization with Lorenz staggering and terrain-following height coordinate.  The implementation includes both fully explicit time integration, and a horizontally-explicit vertically-implicit formulation that is solved with a third-order implicit-explicit additive Runge-Kutta scheme from \cite{ascher1997implicit}.  Fourth-order hyperviscosity is used in the horizontal to prevent a buildup of energy at the grid scale (\S \ref{sec:diffusion_acme}).  The model further provides an optional upwind-biased transport scheme in the vertical column.  Tracer transport is performed using the spectral element method with the same timestep as the hydrodynamics and using the tracer mass density as a prognostic variable.  As with the hydrodynamics, tracer transport is performed explicitly in the horizontal and implicitly in the vertical.

@@

The classic latitude-longitude grid is produced by subdividing the sphere along lines of constant latitude and longitude.  The latitude-longitude grid has the benefits of being globally rectilinear, which simplifies data access and subdivision of computation across processors, and yields a vector basis that is locally orthogonal nearly everywhere.  This structure accurately maintains purely zonal flows and simplifies data post-processing for visualization.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is presently employed in many global models, including the UK Met Office New Dynamics and ENDGame dynamical cores \citep{davies2005new, wood2014inherently}.  The latitude-longitude grid is also an option in the GEM model.

@@

The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed, PL2007JCP} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.  The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.  Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.  This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).  Among the DCMIP2016 models, the cubed-sphere grid is employed by the ACME, FV$^3$, and TEMPEST dynamical cores.

\begin{figure}
\begin{center}
\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{Figure1a.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{Figure1b.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{Figure1c.pdf}
\end{center}
\caption{(Left) A cubed-sphere grid.  (Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.  (Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
\end{figure}

@@

The icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.  These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained. For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle. By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.  Among the DCMIP2016 models, the icosahedral (triangular) grid is employed operationally in the ICON dynamical core.

@@

\subsubsection{Mass-based coordinates} \label{sec:mass_coords}

@@

\subsection{CSU} \label{sec:CSUEquations}
